---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.path = "figure/", dev = "svglite"
)
```

# sugrrants

[![Travis-CI Build Status](https://travis-ci.org/earowang/sugrrants.svg?branch=master)](https://travis-ci.org/earowang/sugrrants)

## Overview

The goal of *sugrrants* is to provide supporting graphics with R for analysing time series data. It aims to fit into the *tidyverse* and grammar of graphics framework for handling with temporal data. The package is under development and highly experimental at this stage. Bug reporting and feature suggestions are welcome.

## Installation

You could install the development version from Github using

```r
# install.packages("devtools")
devtools::install_github("earowang/sugrrants", build_vignettes = TRUE)
```

## Usage

### Calendar-based graphics

```{r calendar, echo = TRUE}
library(dplyr)
library(sugrrants)

calendar_df <- pedestrian %>%
  filter(Sensor_ID %in% c(9, 25)) %>%
  group_by(Sensor_Name) %>%
  frame_calendar(
    x = Time, y = Hourly_Counts, date = Date_Time, calendar = "monthly")
calendar_df
```

```{r calendar-plot, echo = TRUE, fig.height = 8}
p <- calendar_df %>%
  ggplot(aes(x = .x, y = .y, group = .group_id)) +
  geom_line() +
  facet_grid(Sensor_Name ~ .)
prettify(p, label = "text")
```

## Acknowledgements

The acronym of *sugrrants* is **SU**pporting **GR**aphics with **R** for **AN**alysing **T**ime **S**eries, pronounced as "sugar ants" that are a species of ant endemic to Australia. Naming credit goes to @robjhyndman.
