---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.path = "man/figure/"
)
options(tibble.print_min = 5)
```

# sugrrants

[![Travis-CI Build Status](https://travis-ci.org/earowang/sugrrants.svg?branch=master)](https://travis-ci.org/earowang/sugrrants)
[![Coverage Status](https://img.shields.io/codecov/c/github/earowang/sugrrants/master.svg)](https://codecov.io/github/earowang/sugrrants?branch=master)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/sugrrants)](https://cran.r-project.org/package=sugrrants)
[![Downloads](http://cranlogs.r-pkg.org/badges/sugrrants?color=brightgreen)](https://cran.r-project.org/package=sugrrants)

The goal of *sugrrants* is to provide supporting graphs with R for analysing time series data. It aims to fit into the *tidyverse* and grammar of graphics framework for handling temporal data.

## Installation

You could install the stable version on CRAN:

```{r, eval = FALSE}
install.packages("sugrrants")
```

You could also install the development version from Github using:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("earowang/sugrrants", build_vignettes = TRUE)
```

## Usage

### Calendar-based graphics

```{r facet-calendar, echo = TRUE, message = FALSE, fig.height = 9, fig.width = 9}
library(dplyr)
library(sugrrants)
pedestrian %>%
  filter(Date < as.Date("2016-05-01")) %>% 
  ggplot(aes(x = Time, y = Hourly_Counts, colour = Sensor_Name)) +
  geom_line() +
  facet_calendar(date = Date) +
  theme(legend.position = "bottom")
```

The `frame_calendar()` provides tools for re-structuring the data into a calendar layout, without using the faceting method. It is fast, compact and light-weight, although it does not preserve the values.

```{r calendar-plot, echo = TRUE, fig.height = 7, fig.width = 7}
p <- pedestrian %>%
  filter(Sensor_ID == 9, Year == 2016) %>%
  mutate(Weekend = if_else(Day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")) %>%
  frame_calendar(x = Time, y = Hourly_Counts, date = Date) %>% 
  ggplot(aes(x = .Time, y = .Hourly_Counts, group = Date, colour = Weekend)) +
  geom_line() +
  theme(legend.position = "bottom")
prettify(p)
```

## Google Summer of Code 2017

This package is part of the project---[Tidy data structures and visual methods to support exploration of big temporal-context data](https://summerofcode.withgoogle.com/projects/#4790455121215488), which has been participated in Google Summer of Code 2017 (gsoc), for [R project for statistical computing](https://www.r-project.org). 

A new function `frame_calendar()` [[here](https://github.com/earowang/sugrrants/blob/master/R/frame-calendar.R) and [here](https://github.com/earowang/sugrrants/blob/master/R/calendar-fun.R)] in the **sugrrants** package has been developed and documented for calendar-based graphics. I have also written a vignette [[source](https://github.com/earowang/sugrrants/blob/master/vignettes/frame-calendar.Rmd) and [reader view](http://pkg.earo.me/sugrrants/articles/frame-calendar.html)], which introduces and demonstrates the usage of the `frame_calendar()` function. [Many unit tests](https://github.com/earowang/sugrrants/blob/master/tests/testthat/test-calendar.R) have been carried out to ensure the expected performance of this function. The function implements non-standard evaluation and highlights the [tidy evaluation](http://rlang.tidyverse.org/articles/tidy-evaluation.html) in action. The initial release (v0.1.0) of the package has been published on [CRAN](https://CRAN.R-project.org/package=sugrrants) during the gsoc summer time.

I have initialised a new R package [**tsibble**](https://github.com/earowang/tsibble) for tidy temporal data, as part of the project. The `tsibble()` function constructs a new `tbl_ts` class for temporal data, and the `as_tsibble()` helps to convert a few `ts` objects into the `tbl_ts` class. Some key verbs (generics) from the **dplyr** package, such as `mutate()`, `summarise()`, `filter()`, have been defined and developed for the `tbl_ts` data class. The **tsibble** package was highly experimental over the period of the gsoc [[commits](https://github.com/earowang/tsibble/commit/aba1cfc2eec88966c43232fe5d249522f88e1e27)], and these functions are very likely to be changed or improved in the future.

A new package [**rwalkr**](https://github.com/earowang/rwalkr) has been created and released on [CRAN](https://cran.r-project.org/package=rwalkr) during the gsoc summer. This package provides API to Melbourne pedestrian sensor data and arrange the data in tidy temporal data form. Two functions including [`walk_melb()`](https://github.com/earowang/rwalkr/blob/master/R/scrape.R) and [`shine_melb()`](https://github.com/earowang/rwalkr/blob/master/R/shiny.R), have been written and documented as the v0.1.0 and v0.2.0 releases on CRAN. The majority of the code for the function [`run_melb()`](https://github.com/earowang/rwalkr/blob/master/R/soda.R) has been done, but the interface needs improving after the gsoc.

## Miscellaneous

The acronym of *sugrrants* is **SU**pporting **GR**aphs with **R** for **AN**alysing **T**ime **S**eries, pronounced as "sugar ants" that are a species of ant endemic to Australia.

---
Please note that this project is released with a [Contributor Code of Conduct](.github/CODE_OF_CONDUCT.md). By participating in this project you agree to abide by its terms.
