---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.path = "figure/"
)
```

# sugrrants

[![Travis-CI Build Status](https://travis-ci.org/earowang/sugrrants.svg?branch=master)](https://travis-ci.org/earowang/sugrrants)

## Overview

The goal of *sugrrants* is to provide supporting graphics with R for analysing time series data. It aims to fit into the *tidyverse* and grammar of graphics framework for handling with temporal data. The package is under development and highly experimental at this stage. Bug reporting and feature suggestions are welcome.

## Installation

You could install the development version from Github using

```r
# install.packages("devtools")
devtools::install_github("earowang/sugrrants", build_vignettes = TRUE)
```

## Usage

### Calendar-based graphics

```{r calendar, echo = TRUE, message = FALSE}
library(dplyr)
library(sugrrants)

calendar_df <- pedestrian %>%
  filter(Sensor_ID == 9, Year == 2016) %>%
  mutate(
    Weekend = if_else(Day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
  ) %>%
  frame_calendar(
    x = Time, y = Hourly_Counts, date = Date, calendar = "monthly"
  )
calendar_df
```

```{r calendar-plot, echo = TRUE, fig.height = 7, fig.width = 7}
p <- calendar_df %>%
  ggplot(aes(x = .Time, y = .Hourly_Counts, group = Date, colour = Weekend)) +
  geom_line() +
  theme(legend.position = "none")
prettify(p)
```

## Miscellaneous

The acronym of *sugrrants* is **SU**pporting **GR**aphics with **R** for **AN**alysing **T**ime **S**eries, pronounced as "sugar ants" that are a species of ant endemic to Australia.
